---
title: Epidemiological Trends and Prevalence of Cleft Lip and Palate in Bhutan
author: "Tshewang Gyeltshen, Karma Tobgay"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
    number_sections: no
    toc_depth: 3
    df_print: paged
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  chunk_output_type: console
  theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="asis",warning = FALSE, message = FALSE)
rm(list = ls())
setwd("D:/Research/Bhutan_cleft_epi/")
```

#Loading the required library packages

```{r Loading library, echo=FALSE}
library(dplyr)
library(ggplot2)
library(arsenal)
library(knitr)
library(gtsummary)
library(ggplot2)
library(Hmisc)

```

#Load and Prepare Data
```{r Loading and data preparation}
#Loading the dataset
cleft <- readxl::read_xlsx("D:/Research/Bhutan_cleft_epi/NBD_data2015_20222.xlsx")

#Selecting the required variables for the study 
cleft001 <- cleft %>%
  select(
    hosp_name = `Hospital Name`,
    birthdate = `Baby Birth Date`,
    m_age_cat = `Mom's Age`,
    m_age = `Mother Age`,
    f_age_cat = `Dad's Age`,
    f_age = `Father Age`,
    sex = `Baby Gender`,
    birthwgt_cat = `Birth-Weight`,
    birthwgt = `Baby Birth Weight`,
    hd_crcmfrnce = `Head Circumference`,
    gestage = `Gestational-Age`,
    gestwk = `Gestation Week`,
    #icd = `ICD Code`,
    bd_cat_main = `Main BD Category`,
    bd_cat_sub = `Sub BD Category`
  ) %>%
  filter(bd_cat_main == "Q35-Q37 Cleft lip and palate") %>% 
  mutate(
    year = as.numeric(format(birthdate, '%Y')),  # extracting year from the baby birth-date
    hosp_name = case_when(
        hosp_name == "ERRH: Eastern Regional Referral Hospital, Mongar" ~ "ERRH",
        hosp_name == "JDWNR: Jigme Dorji Wangchuck National Referral Hospital" ~ "JDWNRH",
        TRUE ~ hosp_name ),
    hd_crcmfrnce = ifelse(hd_crcmfrnce == -1, NA, hd_crcmfrnce), 
         f_age = if_else(f_age == -1, NA, f_age)
    )

pop_btn <- data.frame(year = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022), 
                      male_pop = c(393324, 398948, 380453, 383909, 387402, 390877, 394323, 397731), 
                      female_pop = c(363718, 369629, 346692, 350465, 354270, 358054, 361806, 365518), 
                      lb_male = c(NA, 5400, 5423, 5753, 5191, 5266, 5452, NA), 
                      lb_female = c(NA, 5057, 5189, 5615, 4995, 4974, 5303, NA), 
                      cbr = c(17.5, 14.4, 15.5, 15.5, 13.7, 13.7, 14.2, 16.00 )) #2017&2022-Bhu_Pop_Proj.pdf
#vital statistics 2021 & 2022 NSB
pop_btn$total <- pop_btn$male + pop_btn$female
# Srce: Population Projections of Bhutan, 2005-2030, and 2017-2047; NSB

pop_btn <- pop_btn %>% 
  mutate(cbr_n = ((lb_male+lb_female)/total)*1000,
         cbr_n = if_else(is.na(cbr_n), cbr, cbr_n))

tot_defect <- cleft %>%
   mutate(
    year = as.numeric(format(`Baby Birth Date`, '%Y'))) %>% 
  group_by(year) %>%
  summarise(
    total_defect = n())

cleft002 <- cleft001 %>%
  mutate(f_age = if_else(f_age == -1, NA_real_, f_age), 
         hd_crcmfrnce = if_else(hd_crcmfrnce == -1, NA_real_, hd_crcmfrnce)) %>% 
  group_by(year) %>%
  summarise(
    obs = n(),
    m_age = mean(m_age),
    #f_age = mean(f_age, na.rm = T),
    hd_crcmfrnce = mean(hd_crcmfrnce, na.rm = T),
    sexm = sum(sex =="Male"),
    sexf = sum(sex == "Female"),
    birthwgt = mean(birthwgt),
    gestwk = mean(gestwk)
  )

cleft002 <- within(cleft002, {
  femalepop <- pop_btn$female
  malepop <- pop_btn$male
  totalpop <- femalepop + malepop
  lb_male <- pop_btn$lb_male
  lb_female <- pop_btn$lb_female
  cbr <- round(pop_btn$cbr_n, digits = 2)
  total_birth <- pop_btn$lb_male + pop_btn$lb_female
  total_birth <- if_else(is.na(total_birth), 
                         round(cbr*totalpop/1000, digits = 0), 
                         total_birth)
  total_defect <- tot_defect$total_defect
  #year <- factor(year)
})

cleft002 <- cleft002 %>% 
  select(year, obs, total_defect, lb_female, lb_male, total_birth, cbr, femalepop, malepop, totalpop, sexm, sexf, m_age, hd_crcmfrnce, birthwgt, gestwk )

cleft002$clp_pertge <- round(cleft002$obs/cleft002$total_birth*1000, digits = 2)

cleft002$dft_pertge <- round(cleft002$total_defect/cleft002$total_birth*1000, digits = 2)

#save(cleft002, file ="cleft002.rda")

```

# Descriptive Statistics
Running Descriptive Statistics and exploring dataset
```{r Descriptive Statistics}
# summarizing by specific defects
summary(freqlist(~bd_cat_sub, labelTranslations = 
                   c(bd_cat_sub = "Specific Birth Defect Category"),cleft001, na.options = "remove")
        )
```

## Table 1. Characterstics of the Babies

    -   Year
    -   Sex of the Babies (sex)
    -   Age of the mother (m_age_cat)
    -   Age of the father (f_age_cat)
    -   Birth Weight Category (birthwgt_cat)
    -   Head Circumference
    -   Gestational Age Category (gestage)
    
```{r tab1_Demographic_characterstics}
tab1 <- tableby( ~  factor(year) + sex + m_age_cat + f_age_cat + birthwgt_cat + hd_crcmfrnce + gestage, data= cleft001, test=F)
labels(tab1)  <- c(`factor(year)` = "Year", 
                   sex = "Sex of the Babies", 
                   m_age_cat= "Age Category of the Mother", 
                   f_age_cat= "Age Category of the Father", 
                   birthwgt_cat = "Birth Weight Category", 
                   gestage = "Gestational Age Category", 
                   hd_crcmfrnce = "Head Circumference"
                   )
summary(tab1, title='**Table 1: Characterstics of the cleft lip and palate babies**', pfootnote = TRUE)

#write2word(tab1, "Characterstics.doc", title = "Table 1.Characteristics of the survey participants")

```

# Trends and Visualization
```{r trend_graph}
#Trend over the years 
fig1<- cleft001 %>%
  group_by(year) %>%
  summarise(obs = n()) %>%
  ggplot(aes(year, obs)) + 
  geom_point(shape = 20, colour = "blue", size = 5) + 
  geom_text(aes(label = obs), vjust = -0.5) +  # Label the count points
  labs(title = "Prevalence", 
       x = "Years", 
       y = "Number of Cleft Babies born") +
  geom_smooth(method = NULL, aes(color = "red"), se = F) + 
  scale_x_continuous(breaks = unique(cleft001$year)) +  # Ensure all years are displayed
  theme_classic() +
   theme(plot.title = element_text(hjust = 0.5))

#Oro-facial Anomalies over the years 
count <- cleft001 %>% 
  filter(!is.na(bd_cat_sub)) %>%
  group_by(year) %>%
  count(bd_cat_sub)

ggplot(count, aes(x = year, y = n, fill = bd_cat_sub, colour = bd_cat_sub)) + 
  geom_bar(stat = "identity", position = "dodge", alpha = 0.5) + 
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, 
            size = 3, 
            show.legend = FALSE)+
  labs(title = "Total Cleft Babies by specific cleft type in Bhutan, 2015 - 2022", 
       x = "Years", 
       y = "Number of infants with cleft lip or palate", 
       fill = "", 
       color = "") + # removes the variable name from the legend
  theme_classic() +
  theme(legend.position = "bottom",  # Position legend at the bottom
        legend.text = element_text(size = 9)) +  # Adjust legend text size
  scale_x_continuous(breaks = unique(cleft001$year)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = NULL, colour = NULL, name = NULL)


#Visualizing correlation between head circumference & Baby weight 
cleft001 %>% 
  ggplot(., 
         aes(birthwgt, hd_crcmfrnce))+ 
  geom_point(shape = 21, colour = "red", size = ) + 
  labs(title = "Correlation between Baby Weight and Head Circumference", 
       x = "Baby Weight in grams", 
       y = "Head Circumference", 
       caption = "Epidemiology of Cleft lip and Palate in Bhutan, 2015-2022")+
  theme_classic() 

#Visualizing correlation between Baby Weight and Gestational Age  
cleft001 %>% 
  ggplot(., 
         aes(gestwk, birthwgt))+ 
  geom_point(shape = 21, colour = "blue", size = 1.5) + 
  labs(title = "Correlation between Gestational Week and Baby Weight", 
       x = "Gestational Week in Weeks time", 
       y = "Baby Weight in grams", 
       caption = "Epidemiology of Cleft lip and Palate in Bhutan, 2015-2022")+
  theme_classic()
library(ggplot2)
library(gridExtra)

# Create a dataframe for IRR values and their 95% CIs
irr_data <- data.frame(
  IRR = c(1.20, 1.60, 1.70, 2.00, 1.60, 1.60, 1.50),
  CI_lower = c(0.52, 0.74, 0.79, 0.96, 0.74, 0.74, 0.68),
  CI_upper = c(2.84, 3.65, 3.85, 4.46, 3.65, 3.65, 3.45), 
  year <- seq(2016, 2022)
)

# Plot the IRR values with error bars representing 95% CIs
fig2 <- ggplot(irr_data, aes(x = factor(year), y = IRR)) +
  geom_point(shape = 20, colour = "red", size = 5) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.1) +
  geom_text(aes(label = format(IRR, nsmall = 2)), vjust = 0, hjust = -0.2) + 
  labs(title = "Incidence",
       x = "Years",
       y = "Incidence Rate Ratio (IRR)") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(fig1, fig2, nrow = 1, top = "Prevalence and Incidence of Cleft Babies in Bhutan, 2015 - 2022")

```

# Correlates and Association
## Table 2. Cross Tabulations by year 

    -   Year
    -   Sex of the Babies (sex)
    -   Age of the mother (m_age_cat)
    -   Age of the father (f_age_cat)
    -   Birth Weight Category (birthwgt_cat)
    -   Head Circumference
    -   Gestational Age Category (gestage)
    
```{r tab2_cross_tabulations}
tab2 <- tableby(year ~  gestage + sex + m_age_cat + f_age_cat + birthwgt_cat + hd_crcmfrnce, data= cleft001, test=T)
labels(tab2)  <- c(gestage = "Gestational Age Category", 
                   sex = "Sex of the Babies", 
                   m_age_cat= "Mother's Age Category", 
                   f_age_cat= "Father's Age Category", 
                   birthwgt_cat = "Birthweight of the Babies", 
                   hd_crcmfrnce = "Head Circumference at Birth"
                   )
summary(tab2, title='**Table 2: Maternal and Child Characteristics**', pfootnote = TRUE)

#write2word(tab2, "M_Characteristics.doc", title = "Table 2.Maternal and Child Characteristics")

```

## Fitting Poisson
```{r, GLM and Model Fitting}
#poisson model to compute the incidence
model <- glm(obs ~ factor(year), family = poisson, data = cleft002)
tbl_regression(model, exponentiate = TRUE)

```
